Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-17 paper/main.tex
comments in section 3

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2024dpc}, we consider two functions 
($f_1$ and $f_2$) each with two variance specifications (settings ``A" and ``B", 
\textcolor{magenta}{we can consider additional homoskedastic cases in the appendix if necessary}).

Here, $f_1$ and $f_2$ are defined as follows:
\begin{align}
  f_1(x) &= m_1 \exp(-u_1x/2) * \cos(wx) - m_1x/5, \quad w=\sqrt{25-[u_1/2]^2} \\
  f_2(x) &= \exp(-m_2[x-3]^2)+\exp(-u_2[x-1]^2)-0.05\sin(8[x-1.9]), \\ 
  \quad x &\in [0,4].
\end{align}

For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are random where 
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$.

For the two variance settings, we consider matrices $\Sigma_A$ and $\Sigma_B$ to 
have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. 
For setting ``A", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``B", we have 
$\tau^2=0.1$ and $\theta=0.05$, and additionally scale this covariance matrix by 
a vector of precision values $s$ so that $\Sigma_B = s M s^T$ is nonstationary; here $M$ is 
a Mat\'ern covariance matrix with parameters $\tau^2$ and $\theta$.

For each simulated function and variance specification, we generate 20 values of 
$m_1, m_2, f_1, f_2$, and conditional on these values, we sample either 5 or 15 
functional observations and estimate the underlying true function based on the observations.

\begin{figure}[t]
    \centering
    \includegraphics[width=6in]{sims_logS.png}
    \caption{Boxplot of log scores across two different functions and covariance 
             settings (lower is better). For each case, the boxplot is constructed 
             from 20 random batches were simulated. Each column represents a function/variance 
             specification pair, and each row shows results for batch sizes of either 
             5 (top) or 15 (bottom).}
    \label{fig:sims_logS}
\end{figure}

The competing methods we consider are an out-of-the-box DGP fit using the 
\texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP fit using 
the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, as well 
as a deep process convolution (DPC) approach, which is utilized within Cosmic 
Emu \citep{moran2023mira}.

The metrics we use to evaluate performance are mean squared error (MSE) and the 
log score, a proper scoring rule which takes both prediction and intervals into 
consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates 
better performance. Log score are shown in Figure \ref{fig:sims_logS}; our DGP.FCO 
model performs favorably against each of these current benchmark methods by
accounting for spatial dependence within our model, resulting in more accurate UQ.
MSE results, which do not address UQ, were comparable across competing methods; these results
can be found in Figure \ref{fig:sims_MSE}. \textcolor{magenta}{If we mention that DGP.FCO timings 
speed up compared to \texttt{deepgp} and DPC, we would also need to mention that \texttt{hetGP} 
is much faster than all of these.}

@@

In addition to simulated data, we investigate model fits for different cosmological 
datasets from CAMB \citep{lewis2011CAMB}. These datasets contain power spectra based
on six-dimensional cosmological parameter settings. Similar to the Mira-Titan suite, the
CAMB dataset has multiple (sixty-four) batches of power spectra for each cosmology, with each batch 
containing multiple (fifteen) low-resolution spectra, and one high-resolution sprectum. 
Unlike the Mira-Titan suite, this dataset favorably contains an infinite-resolution 
spectrum, which can be treated as the true spectrum for each cosmology,
allowing us to accurately assess posterior mean estimates and credible intervals.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{CAMB_fit_model1.png}
    \caption{Example of DGP.FCO model fit for the first batch of model runs from the CAMB data, 
             with LOESS mean term subtracted. We can compare the orange lines 
             (posterior mean and dotted 95\% credible interval) to the light blue line,
             representing the infinite-resolution ``true" spectrum.}   
    \label{fig:fit_camb}
\end{figure}

We use 32 CAMB batches to train the DGP.FCO model and reserve the other 32 as a test set.
To imitate the role of the perturbation theory output of Mira-Titan, we anchor estimates of the CAMB
power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and then rely 
strictly on the low- and high-resolution runs for $k$ above this. We take a weighted
average of the CAMB runs and assume heteroskedastic errors in a similar fashion.

We use a log-log regression model to obtain precision estimates $\Lambda_\ell$ and $\Lambda_h$
for the low- and high-resolution runs respectively, as outlined in Section \ref{sec:data}. 
Similarly, we set $\Lambda_i=10^8$ to illustrate high confidence in the infinite-resolution 
run for low $k$ values and incorporate these estimates within $\Sigma_\varepsilon$. 
With the precision information, we then estimate the Mat\'ern parameters for 
$\Sigma_\varepsilon$. For the CAMB data, lengthscale estimates are consistently 
around $3\times 10^{-7}$, evidence that we can simplify our covariance structure 
to contain no spatial dependence (compare low-resolution runs of Figure \ref{fig:fit_camb} 
with Figure \ref{fig:plot_data}). We therefore specify $\Sigma_\varepsilon = 
(15\Lambda_\ell+\Lambda_h+\Lambda_i)^{-1}$ for the CAMB data.

@@

@@

From plots of the multiple low-resolution spectra (e.g., Figure \ref{fig:plot_data}), 
we find they vary smoothly about some mean $S$, suggesting that a dense covariance 
matrix $\Sigma_\varepsilon$ (accounting for spatial dependence) may outperform a 
diagonal covariance $\Lambda_\ell^{-1}$ for $\bar Y$. From Section \ref{sec:data},
we can incorporate domain knowledge from $\Lambda_p, \Lambda_\ell,$ and $\Lambda_h$ 
within our model. In previous work, we have modeled multiple candidates for 
$\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance 
$\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision 
values and subtracting a LOESS-smoothed average) performs the best. 
We specify this structure within $\Sigma_\varepsilon$ and refer readers 
to \cite{walsh2023bayesian} for a more detailed exposition. This yields 
$\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$, 
where $\Lambda_p, \Sigma_\ell^{-1}, \text{ and } \Lambda_h$ are all $n\times n$ 
and are 0 wherever the corresponding data type is biased. From here, we estimate 
$\tau_\ell, \theta_\ell$ with maximum likelihood, and input these within 
$\Sigma_\varepsilon$, which we use as an input to our model.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.png}
    \caption{95\% credible intervals for the power spectrum of model 1 (posterior mean 
             subtracted). Plots of the perturbation theory, low-resolution and high-resolution 
             runs are also shown, alongside the weighted average.}
    \label{fig:plot_fit}
\end{figure}

@@

>>>>

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2024dpc}, we consider two test functions 
defined as
\[
\begin{aligned}
  f_1(x) &= m_1 \exp\left(-\frac{u_1x}{2}\right) * \cos\left(x\sqrt{25-\left(\frac{u_1}{2}\right)^2}\right) - \frac{m_1x}{5}\\
  f_2(x) &= \exp\left(-m_2(x-3)^2\right)+\exp\left(-u_2(x-1)^2\right)-0.05\sin\left(8(x-1.9)\right.
\end{aligned}
\]
We consider $x\in[0, 4]$.  For additional variation, the values of $m_1, m_2, u_1,$ and 
$u_2$ are randomly selected from
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$.

For the two variance settings, we consider matrices $\Sigma_A$ and $\Sigma_B$ to 
have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. 
For setting ``A", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``B", we have 
$\tau^2=0.1$ and $\theta=0.05$, and additionally scale this covariance matrix by 
a vector of precision values $s$ so that $\Sigma_B = s M s^T$ is nonstationary; here $M$ is 
a Mat\'ern covariance matrix with parameters $\tau^2$ and $\theta$.
\textcolor{red}{(@Steve - this whole paragraph needs re-write.  Need to start saying
how you draw $r$-many realizations from a GP with some mean (?) and some spatially varying
covariance.  Then use something like this:
\[
\begin{aligned}
\Sigma_A &= 0.0225 * k(||x_i - x_j||^2, 0.01) \\
\Sigma_B &= s^\top Ms \quad\textrm{where}\quad M = 0.1 * k(||x_i - x_j||^2, 0.05)
\quad\textrm{and}\quad s = ???
\end{aligned}
\]
where you say $k$ is defined according to the previous equation.  Also, what is s?}

For each simulated function and variance specification, we consider 20 Monte Carlo repetitions
with re-randomized $m_1, m_2, u_1, u_2$.  We consider $r = 5$ and $r = 15$.
\textcolor{red}{@Steve - what is $n$?  How dense is the grid?  I think we need a visual
of this scenario.  Perhaps two panels, one with an $f_1$ and one with an $f_2$.
Show the true function.  Then show $r = 5$ realizations.  Use points to highlight
the training observations.  That way for the other methods, you can comment on how
they treat the points as all the same, not coming from 5 smooth curves.}

\begin{figure}[t]
    \centering
    \includegraphics[width=6in]{sims_logS.png}
    \caption{Boxplot of log scores (20 repetitions, lower is better) for each 
             simulated scenario.  Each column represents a function/covariance 
             pair, and each row shows results for batch sizes of either 
             5 (top) or 15 (bottom) realizations. \textcolor{red}{(This figure needs udpating.)}}
    \label{fig:sims_logS}
\end{figure}

The competing methods we consider are an out-of-the-box DGP fit using the 
\texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP fit using 
the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, as well 
as a deep process convolution (DPC) approach, which is utilized within Cosmic 
Emu \citep{moran2023mira}. \textcolor{red}{(@Steve - given the new figure, comment
on how these methods view the points as iid noise/heteroskedastic noise/etc.
Also need to comment on how $\Sigma_\epsilon$ is estimated for our method.  And
that there is no pre-processing of the weighted average, right?)}

The metrics we use to evaluate performance are mean squared error (MSE) and the 
log score, a proper scoring rule which takes both prediction and intervals into 
consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates 
better performance. Log score are shown in Figure \ref{fig:sims_logS}; our DGP.FCO 
model performs favorably against each of these current benchmark methods by
accounting for spatial dependence within our model, resulting in more accurate UQ.
MSE results, which do not address UQ, were comparable across competing methods; these results
can be found in Figure \ref{fig:sims_MSE}.
\textcolor{red}{(@Steve - this paragraph needs revising now that the figure is moved
to the appendix.)}

@@

We next investigate model fits to several cosmological 
datasets from CAMB \citep{lewis2011CAMB}. These datasets contain power spectra based
on six-dimensional cosmological parameter settings. Similar to the Mira-Titan suite, the
CAMB dataset has multiple (sixty-four) batches of power spectra for each cosmology, with each batch 
containing multiple (fifteen) low-resolution spectra, and one high-resolution sprectum
\textcolor{red}{(@Steve - do you mean there are 64 cosmologies?)} 
Unlike the Mira-Titan suite, this dataset contains an infinite-resolution 
spectrum, which can be treated as the true spectrum for each cosmology,
allowing us to assess the accuracy of posterior mean estimates and effectiveness
of credible intervals. \textcolor{red}{(@Steve - refer to Figure 3 as a visual.  Comment on the
true power spectrum and the low/high res runs.)}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{CAMB_fit_model1.png}
    \caption{Example of DGP.FCO model fit for the first batch of model runs from the CAMB data, 
             with LOESS mean term subtracted. \textcolor{red}{(LOESS mean of what?  We need
             to officially say how things are centered.)} We can compare the orange lines 
             (posterior mean and dotted 95\% credible interval) to the light blue line,
             representing the infinite-resolution ``true" spectrum.}   
    \label{fig:fit_camb}
\end{figure}

We use 32 CAMB batches to train the DGP.FCO model and reserve the other 32 as a test set.
\textcolor{red}{(@Steve - I'm so confused.  We are not talking about PCA here.  Each
cosmology stands on its own.  We use the low/high res runs for training, and the true
spectrum for testing.)}
To imitate the role of the perturbation theory output of Mira-Titan, we anchor estimates of the CAMB
power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and then rely 
strictly on the low- and high-resolution runs for $k$ above this. 
\textcolor{red}{(Refer to the figure here as well to help explain?)}
We take a weighted
average of the CAMB runs and assume heteroskedastic errors in a similar fashion. 
\textcolor{red}{(Need details on how the weighted average is calculated.  Can
refer to previous section, but need to be specific.)}

To estimate $\Sigma_\epsilon$, we first use a log-log regression model to obtain 
precision estimates $\Lambda_\ell$ and $\Lambda_h$
for the low- and high-resolution runs respectively, as outlined in Section \ref{sec:data}. 
Similarly, we set $\Lambda_i=10^8$ \textcolor{red}{(this notation is no longer right.
Also why switch from $10^5$ to $10^8$?)} 
to illustrate high confidence in the infinite-resolution 
run for low $k$ values and incorporate these estimates within $\Sigma_\varepsilon$. 
With the precision information, we then estimate the Mat\'ern parameters for 
$\Sigma_\varepsilon$. \textcolor{red}{(Need more specifics.  Are we using a $\tau^2$ now?
5/2 still?  Maximum likelihood estimation?)}
For the CAMB data, lengthscale estimates are consistently 
around $3\times 10^{-7}$, evidence that we can simplify our covariance structure 
to contain no spatial dependence (compare low-resolution runs of Figure \ref{fig:fit_camb} 
with Figure \ref{fig:plot_data} \textcolor{red}{ - I don't understand what we are supposed
to see from this comparison.)}). We therefore specify $\Sigma_\varepsilon = 
(15\Lambda_\ell+\Lambda_h+\Lambda_i)^{-1}$ for the CAMB data.

\textcolor{red}{(We need to fix the entire previous paragraph. How about we keep it short
and sweet and say we considered a Matern kernel, but found that the spatial correlation
decayed so quickly ($\theta\rightarrow 0$) that spatial dependence was not necessary.)}

@@

\textcolor{red}{(Where is the MSE/logscore across the 64 batches????  Are we comparing
to other methods?  As it stands, this is barely a "benchmark".  It is more of a "proof
of concept".)}

@@

\textcolor{red}{(@Steve - we've done a lot of edits up to this point.  Can
you revise this text now to fit in with the rest?  Need to refer to how the 
weighted average is calculated.
And the purpose of $\Sigma_\epsilon$.  But this is the first time Mira-Titan
specific details of $\Sigma_\epsilon$ will be mentioned.)}
From plots of the multiple low-resolution spectra (e.g., Figure \ref{fig:plot_data}), 
we find they vary smoothly about some mean $S$, suggesting that a dense covariance 
matrix $\Sigma_\varepsilon$ (accounting for spatial dependence) may outperform a 
diagonal covariance $\Lambda_\ell^{-1}$ for $\bar Y$. From Section \ref{sec:data},
we can incorporate domain knowledge from $\Lambda_p, \Lambda_\ell,$ and $\Lambda_h$ 
within our model. In previous work, we have modeled multiple candidates for 
$\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance 
$\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision 
values and subtracting a LOESS-smoothed average) performs the best. 
We specify this structure within $\Sigma_\varepsilon$ and refer readers 
to \cite{walsh2023bayesian} for a more detailed exposition. This yields 
$\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$, 
where $\Lambda_p, \Sigma_\ell^{-1}, \text{ and } \Lambda_h$ are all $n\times n$ 
and are 0 wherever the corresponding data type is biased. From here, we estimate 
$\tau_\ell, \theta_\ell$ with maximum likelihood, and input these within 
$\Sigma_\varepsilon$, which we use as an input to our model.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.png}
    \caption{95\% credible intervals for the power spectrum of model 1 (posterior mean 
             subtracted). Plots of the perturbation theory, low-resolution and high-resolution 
             runs are also shown, alongside the weighted average.
             \textcolor{red}{(This plot needs to be made fresh (not using the package
             function.  Adjust y-axis.  We don't need the yellow/red coloring.)}}
    \label{fig:plot_fit}
\end{figure}

@@

\textcolor{red}{(This whole subsection is missing a point.  What is the point?
We don't know the truth, but what can we glean from this?  What does the warping
tell us?  Talk about the UQ...)}
